Collaborative Multi-output Gaussian Processes: \\ Supplementary Material

### 2014-06-11 paper/inference.tex
other minor fixes

%\subsection{Variational Inference in Sparse GPs Revisited}
%
%In this section we review inference in sparse GPs (as presented in Titsias).
%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
%
We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ using variational inference.
In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs. Section \ref{sec:svi}  takes advantage of this factorization to 
derive stochastic variational inference, allowing the model to scale to very large datasets.
Section \ref{sec:complexity} compares the complexity of the model with previous multi-output methods.

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
We first observe that the true posterior distribution can be written as:
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
to derive variational inference algorithms for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \eqref{eq:gu} and \eqref{eq:hv}), 
we only need to 
learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
the true posteriors.
The quality of approximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.

\noindent
To find the best distribution $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergence terms are analytically tractable.
To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}

>>>>

%\subsection{Variational Inference in Sparse GPs Revisited}
%
%In this section we review inference in sparse GPs (as presented in Titsias).
%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
%
We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ using variational inference \citep{jordan-variational-99}.
In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs. Section \ref{sec:svi}  takes advantage of this factorization to 
derive stochastic variational inference, allowing the model to scale to very large datasets.
Section \ref{sec:complexity} compares the complexity of the model with previous multi-output methods.

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
In variational inference, we find the "closest" approximate distribution to the true posterior in terms of the $\KL$ divergence.
We first observe that the true posterior distribution can be written as:
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
to derive variational inference algorithms for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \eqref{eq:gu} and \eqref{eq:hv}), 
we only need to 
learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
the true posteriors.
The quality of approximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.

\noindent
To find the "best" $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergence terms are analytically tractable.
To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}

### 2014-06-11 paper/model.tex
other minor fixes

Before diving into technical details of the model specification, 
we discuss the modeling philosophy behind our collaborative multi-output Gaussian processes.
To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
This is achieved in the model by allowing the outputs to share multiple sets of inducing variables, 
each of which captures a different pattern common to the outputs.
These variables  play a double pivotal role in the model: they collaboratively share  information 
across the outputs and  provide \emph{sufficient statistics}  so as to induce sparse processes.

>>>>

Before diving into technical details of the model specification, 
we discuss the modeling philosophy behind our collaborative multi-output Gaussian processes.
To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
This is achieved in the model by allowing the outputs to share multiple sets of inducing variables, 
each of which captures a different pattern common to the outputs.
These variables  play a double pivotal role in the model: they collaboratively share  information 
across the outputs and  provide sufficient statistics so as to induce sparse processes.
